Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2005-06-27 paper/metamonads.tex
minor fixes to english in paper.  add submitted mm.zip file

|Blitz++| \cite{Veldhuizen:1998:ISCOPE} and {C++} template
meta-programming in general similarly eliminate levels
of abstraction.  With traits and concepts, some domain-specific
knowledge can also be encoded.  However overhead elimination
critically depends on the compiler's fully inlining of all methods,
which has been reported to be challenging to insure. Furthermore, all
errors (such as type errors and concept violation errors, i.e.,
composition errors) are detected only when compiling the generated
code. It is immensely difficult to correlate errors (e.g., line
numbers) to the ones in the generator itself.

@@

The paper \cite{GluckJ97} describes early simple experiments in
\emph{automatic} and manual staging, and the multi-level language
based on an annotated subset of Scheme (which is untyped and has no
imperative features). The generated code requires post-processing to
attain efficiency.  We are looking into encapsulating staging
annotations in few functors, so that the rest of the code (in
particular, the |Gen| functor that puts it all together) should be
annotation-free and so can be used as it is in one-stage environment
(pure OCaml) as well as multi-stage environment (generating
extensions). The one-stage code is a good baseline for benchmarks and
regression tests. Obtaining a generating extension from properly
modularized OCaml code (along the lines of our |Gen|) is an exciting
area of our future research.

>>>>

|Blitz++| \cite{Veldhuizen:1998:ISCOPE} and {C++} template
meta-programming in general similarly eliminate levels
of abstraction.  With traits and concepts, some domain-specific
knowledge can also be encoded.  However overhead elimination
critically depends on full inlining of all methods by the compiler,
which has been reported to be challenging to insure. Furthermore, all
errors (such as type errors and concept violation errors, i.e.,
composition errors) are detected only when compiling the generated
code. It is immensely difficult to correlate errors (e.g., line
numbers) to the ones in the generator itself.

@@

The paper \cite{GluckJ97} describes early simple experiments in
\emph{automatic} and manual staging, and the multi-level language
based on an annotated subset of Scheme (which is untyped and has no
imperative features). The generated code requires post-processing to
attain efficiency.

We are looking into encapsulating staging
annotations into just a few functors, so that the rest of the code (in
particular, the |Gen| functor that puts it all together) should be
annotation-free and thus can be used as is in a one-stage environment
(pure OCaml) as well as in a multi-stage environment (generating
extensions). The one-stage code is a good baseline for benchmarks and
regression tests. Obtaining a generating extension from properly
modularized OCaml code (along the lines of our |Gen|) is an exciting
area of our future research.
